Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-09 Paper/paper.tex
polishing s3 more

\newcommand{\ii}{\mathrm{i}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}
\newcommand{\arr}{\mathcal{r}}

@@

% #################
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR}
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\begin{figure*}[!t]
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear002}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.
% We do so by comparing the systematic biases introduced by this 
% assumption to the overall statistical uncertainties in the measurement of binary 
% mass and spin parameters. We find that unless the single-detector SNR of the signal
% exceeds $30$ (for design Advanced LIGO sensitivity), the bias accurued by ignoring
% tidal effects in filtering templates will remain smaller than the statistical 
% uncertainty of the same measurement.

@@

We construct NSBH waveforms using the LEA+ model. We inject these into zero
noise at different SNRs, and run MCMC on the injections using BBH 
templates (LEA+ with tidal terms set to zero). We perform this analysis for mass
ratios $q\in\{2,3,4,5\}$, i.e.
$\mbh\in\{2.7M_\odot, 4.05M_\odot, 5.4M_\odot, 6.75M_\odot\}$, and BH spins
$\chibh\in\{-0.5, 0, +0.5, +0.75\}$. We always use $\mns=1.35M_\odot$ and
$\chins=0$, but explore $\lambdans\in\{500, 800, 1000, 1500, 2000\}$.

@@

Next, to get a better understanding of how {\it precisely} we measure NSBH
binary parameters given BBH waveforms, we show the measured $90\%$ confidence
intervals for binary masses and spins in Fig.~\ref{fig:CIWidths90_Lambda_SNR}.
The three panels, corresponding to $\mchirp$ (top), $\eta$ (middle), and $\chibh$
(bottom), show the width of these confidence intervals as a function of BH
mass/spin (within each sub-panel), and NS properties, i.e. $\lambdans$
(downwards in each column).~\footnote{We restrict NS mass to $1.35M_\odot$ and
its spin to zero. Varying its tidal deformability $\lambdans$ does not change
the measurement uncertainties for non-tidal binary parameters, as is evident
from comparing the two rows in each panel of Fig.~\ref{fig:CIWidths90_Lambda_SNR}.}
% 
From the left-most column, we find that: (i) at $\rho=20, \mchirp$ is already
measured remarkably well - to a precision of $0.16\%$ of its true value, and
(ii) so is $\chibh$. (iii) Mass ratio $\eta$ is determined somewhat worse,
with $\gtrsim 25\%$ uncertainty. If the signal is even louder ($\rho\geq 30$),
all three measurements gain further precision, with $\eta$ errors shrinking down
to single-digit percents of its true value.
% 
However, we remind ourselves that there are physical effects that have not been
incorporated in our (BBH) search templates which we use here to decipher NSBH
signals. Therefore, we cannot know the {\it accuracy} of our parameter
measurements unless we know how large the systematic biases are due to an
incomplete template model. To quantify this, we find it useful to define $\arr$
as the ratio of the systematic biases induced by unmodeled tidal effects in
templates, to the statistical uncertainties associated with the measurement
itself. If $\arr\gtrsim 1$, including the effect of NS's
tidal distortion (during inspiral) and possible disruption (close to merger)
would be imperative to yield reliable estimates of binary masses and spins
from its measured GW signal.

In Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR} we show the ratio of the systematic
bias in the measured binary chirp mass $\mchirp$ versus its true value, to the 
statistical uncertainty of the measurement itself. $\mchirp$ is the mass combination
that occurs at leading order in the perturbative expression for GW strain from 
compact binaries, and is therefore very precisely measurable. Even so, we find that
for $\rho=30$ the signal is not loud enough for the systematic bias in $\mchirp$
to exceed a fraction of the statistical uncertainty. When we increase the SNR to 
$\rho=50$ (middle column), we find that if we have two favorable conditions present:
(a) that the BH spin be $\gtrsim 0.4$, {\it and} (b) the NS's true deformability
be large enough that its $\lambdans \gtrsim 1000$, the systematic biases from
ignoring tidal effects in templates reach between $50-100\%$ of the statistical
uncertainty. Therefore at this signal strength, including terms involving the neutron
star's internal structure in waveform models becomes increasingly gainful. At 
even higher SNRs, say $\rho=70$ (right column), under the same favorable conditions
as listed above, the systematic bias in $\mchirp$ measurement {\it exceeds} its 
statistical measurement uncertainty. If the BH is even mildly spinning, say
$\chibh\gtrsim 0.2$, with the NS mildly deformable, say $\lambdans\gtrsim 800$, 
we still are affected in our $\mchirp$ measurement at the $50\%$ level by the 
use of point-particle waveform templates. Finally, we note that the measured (median)
$\mchirp$ is always higher than its true value. This is reasonable since the tidal 
deformation of NSs close to merger reduces the GW signal power in high frequencies, so
it would look like a BH-BH signal of a higher mass (and therefore lower merger or
peak GW-power frequency).

In the next figure, Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR}, we show the same
information (i.e. the ratio of systematic bias to statistical uncertainty in parameter
measurement) for the symmetric mass-ratio $\eta$. Note that $\eta$
is bounded from above and below to within $\eta\in[0, 1/4]$. We immediately observe
that the measured (median) $\eta$ is always lowered below its true value due to the 
lack of $\lambdans$-dependent terms in the filter template model. This bias, as for 
the $\mchirp$ bias, decreases the merger frequency for the BH-BH template, which must
better match the NS-BH merger with its quietened merger and ringdown. At SNR $\rho=30$
the systematic bias in $\eta$ measurement stays mostly around $10-50\%$ of its statistical
measurement uncertainty. For the sole case of a very deformable NS with $\lambdans=2000$,
the missing tidal pieces in template models could cause errors that are comparable to 
that due to noise (i.e. statistical uncertainties). For a still stronger signal with 
$\rho=50$, we find that 
when the following conditions are fulfilled, the measurement of $\eta$ is {\it completely
compromised} due to template biases: (a) BH mass $\mbh\leq 5M_\odot$, (b) BH spin
$\chibh\gtrsim +0.4$, and (c) $\lambdans\gtrsim 1500$. Even under more moderate restrictions
on BH and NS parameters, we find that $\rho=50$ is loud enough to necessitate the 
use of tidal-modeled templates. This observation is further strengthened if we divert 
our attention to the third column in Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR} that presents
the case for even higher SNRs ($\rho=70$).

>>>>

\newcommand{\ii}{\mathrm{i}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}
\newcommand{\arr}{\mathtt{r}}

@@

% #################
\begin{figure*}
\centering 
\includegraphics[width=1.8\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering    
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\includegraphics[width=1.8\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear002}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=1.8\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
}
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.
% We do so by comparing the systematic biases introduced by this 
% assumption to the overall statistical uncertainties in the measurement of binary 
% mass and spin parameters. We find that unless the single-detector SNR of the signal
% exceeds $30$ (for design Advanced LIGO sensitivity), the bias accurued by ignoring
% tidal effects in filtering templates will remain smaller than the statistical 
% uncertainty of the same measurement.

@@

We construct NSBH waveforms using the LEA+ model. We inject these into zero
noise and run MCMC on the injections using BBH templates (LEA+ with tidal terms
set to zero). We perform this analysis for a variety of mass ratios
$q\in\{2,3,4,5\}$, i.e. $\mbh\in\{2.7M_\odot, 4.05M_\odot, 5.4M_\odot, 6.75M_\odot\}$,
and BH spins $\chibh\in\{-0.5, 0, +0.5, +0.75\}$. We always use $\mns=1.35M_\odot$ and
$\chins=0$, but explore $\lambdans\in\{500, 800, 1000, 1500, 2000\}$. The signal
strength is varied from $\rho=20\rightarrow70$. Our choice of this SNR range is
motivated by previous studies which suggest that signatures of NS tidal effects
become observable only at relatively large SNRs~\cite{Lackey:2013axa}.

@@

Next, to get a better understanding of how {\it precisely} we measure NSBH
binary parameters given BBH waveforms, we show the $90\%$ confidence intervals
associated with NSBH mass and spin measurements in Fig.~\ref{fig:CIWidths90_Lambda_SNR}.
The three panels, corresponding to $\mchirp$ (top), $\eta$ (middle), and $\chibh$
(bottom), show the width of these confidence intervals as a function of BH
mass/spin (within each sub-panel), and NS properties, i.e. $\lambdans$
(downwards in each column)~\footnote{We restrict NS mass to $1.35M_\odot$ and
its spin to zero. Varying its tidal deformability $\lambdans$ does not change
the measurement uncertainties for non-tidal binary parameters, as is evident
from comparing the two rows in each panel of Fig.~\ref{fig:CIWidths90_Lambda_SNR}.}.
% 
From the left-most column, we find that: (i) at $\rho=20, \mchirp$ is already
measured remarkably well - to a precision of $0.16\%$ of its true value, and
(ii) so is $\chibh$. (iii) Mass ratio $\eta$ is determined more loosely,
with $\gtrsim 25\%$ uncertainty. If the signal is even louder ($\rho\geq 30$),
all three measurements gain further precision, with $\eta$ errors shrinking down
to single-digit percents of its true value.
% 
However, we remind ourselves that there are physical effects that have not been
incorporated in our (BBH) search templates which we use here to decipher NSBH
signals. Therefore, we cannot know the {\it accuracy} of our parameter
measurements unless we know how large the systematic biases due to an
incomplete template model are. To quantify this, we find it useful to define
$\arr$, the ratio of the systematic biases induced by unmodeled tidal effects in
templates, to the statistical uncertainties associated with the measurement
itself. If $\arr\gtrsim 1$, including the effect of NS's tidal distortion
(during inspiral) and possible disruption (close to merger) would be imperative
to obtain {\it accurate} estimates of source masses and spins from observed GW
signals.

We show this ratio of errors $\arr$ associated with chirp mass measurements in
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}. $\mchirp$ is the mass combination
that occurs at leading order in the perturbative expansion of GW strain emitted
by compact binaries as they spiral in, and is therefore determined the most
precisely. From the left-most column of the figure, we find that even at $\rho=30$, 
the systematic bias in $\mchirp$ measurement is much smaller than its statistical
uncertainty. For signals with $\rho\leq 30$, therefore, we are justified in using
BBH templates to estimate binary chirp mass from NSBH signals. For louder and less
realistic SNRs (say $\rho=50$), we find that $\arr$ can become comparable to unity
if additionally: (i) the BH has prograde spin $\chibh\gtrsim 0.4$, {\it and} (ii)
the NS's true deformability is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
At even higher SNRs, say $\rho=70$, the ratio of errors $\arr$ is large
($\simeq 0.5-1.5$) across most of the binary parameter space; and $\mchirp$ measurement
is affected at a $50\%$ or higher level by the use of point-particle waveforms as
templates.
% 
We conclude that only for $\rho\gtrsim 50$ do we need to include in our templates
terms that capture the effect of neutron star's structural dynamics on emitted GWs.
For lower SNRs, improvements in template model will be washed out by the statistical
measurement uncertainty for $\mchirp$.
% 
As a last note, we find that $\arr$ is always positive for $\mchirp$, suggesting that
it is always over-estimated.
This is understandable since the tidal deformation of NSs close to merger reduces
the GW signal power at high frequencies, which would make the resulting signal
resemble a BBH signal of higher mass (and therefore lower merger frequency).

In the next figure, Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR}, we show the ratio of
measurement errors $\arr$ for the symmetric mass-ratio $\eta$. Note that $\eta$
is bounded from above and below to within $\eta\in[0, 1/4]$. We immediately observe
that the measured (median) $\eta$ is always lowered below its true value due to the 
lack of $\lambdans$-dependent terms in the filter template model. This bias,
similar to the $\mchirp$ bias, decreases the merger frequency for BBH templates,
in order to better match the relatively quiet merger and ringdown of NSBHs.
At SNR $\rho=30$
the systematic bias in $\eta$ measurement stays mostly around $10-50\%$ of its statistical
measurement uncertainty. For the sole case of a very deformable NS with $\lambdans=2000$,
the missing tidal pieces in template models could cause errors that are comparable to 
that due to noise (i.e. statistical uncertainties). For a still stronger signal with 
$\rho=50$, we find that 
when the following conditions are fulfilled, the measurement of $\eta$ is {\it completely
compromised} due to template biases: (a) BH mass $\mbh\leq 5M_\odot$, (b) BH spin
$\chibh\gtrsim +0.4$, and (c) $\lambdans\gtrsim 1500$. Even under more moderate restrictions
on BH and NS parameters, we find that $\rho=50$ is loud enough to necessitate the 
use of tidal-modeled templates. This observation is further strengthened if we divert 
our attention to the third column in Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR} that presents
the case for even higher SNRs ($\rho=70$).
